---
title: "Efforts in Estimating the Incubation Period of 2019-nCoV"
author: "Nianqiao Ju"
date: "2/14/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(devtools)
library(dplyr)
library(ggplot2)
devtools::load_all()
```

## Data Collection
We have collected detailed information about infections in China.
```{r import data}
rm(list = ls())
data <- read.table("../data-raw/Feb14 - In China.tsv", sep = "\t", header = TRUE)
# cases.in.china <- data[c(1:161, 208:449), ]
# use_data(cases.in.china, overwrite = TRUE)
```

We convert the dates to 'days since 1-Dec-2019' format. In order to estimate the incubation period, we only include records with known symptom onset date or period. We assume that 'symptom onset date' is missing completely at random. Currently, the `parse.infected' function assumes Arrival > Symptom. Which is violated by ~ 32 cases.
```{r process data}
## convert the dates to days since 1-Dec-2019
data$Confirmed <- date.process(data$Confirmed)
data$Arrived <- date.process(data$Arrived)
data$Symptom <- date.process(data$Symptom)
data$Initial <- date.process(data$Initial)
data$Hospital <- date.process(data$Hospital)
data$Source <- data$Cluster <- data$Death <- data$Discharged <- data$Verified <- data$Note <- NULL ## for visualization
## parse data
case72 <- data[72,] ## manually parse hefei-72
data <- data[-72,]
data <- parse.infected(data)
case72$Infected_first <- date.process("15-Jan")
case72$Infected_last <- date.process("21-Jan")
data <- rbind(data, case72)
```

The function parse.infection might problematic, because there are 6 cases whose imputed first infection date is greater than last infection date. 
```{r error}
print(data[which(data$Infected_first > data$Infected_last),])
data <- data %>% subset(Infected_first <= Infected_last)
```


We denote by $T_i > 0$ the infection period of each patient, the time from infection to symptom onset. Confirmed cases who had not shown symptoms at the time of confirmation are excluded in this analysis. Infection date information is available for two types of patients: 

* Patients who left Wuhan before 23-Jan-2020 and showed symptoms after leaving Wuhan. We have an lower bound $L_i \leq T_i$ for these patients.

* Some secondary infections results from interacting with patients from Hubei province. Some of these interactions, for example reunion with schoolmates or visiting a shopping center, are observed to the precision of one day or a few days. We observe $L_i \leq T_i \leq U_i$.

* Most of the secondary infections were due to living with family members from outside and their infection date must be later than the arrival date of the family member. As a result, an upper bound can be imputed $0 < T_i \leq U_i.$ For those families that self-quarantined, we can impute also a lower bound $L_i < T_i < U_i$.

```{r subset data}
## Only consider cases with known symptom onset
data <- subset(data, !is.na(Symptom))
## we may choose only to study patients that left Wuhan on or before January 23
# data <- subset(data, Arrived <= 23+31)
## remove outside cases without Ui
data <- subset(data, !(is.na(Arrived) & Infected_first == 1)) 
## this is what we have earlier, it considers only the cases with both bounds
print(dim(data))
```
```{r include=FALSE}
upper <- data %>% subset(Infected_last == Symptom & Infected_first != 1 & Infected_first < Infected_last) %>% mutate(upper = Symptom - Infected_first)
lower <- data %>% subset(Infected_first == 1) %>% mutate(lower = Symptom - Infected_last)
exact <- data %>% subset(Infected_first == Infected_last) %>% mutate(incub = Symptom - Infected_first)
band <- data %>% subset( (Infected_first != 1) & (Infected_first < Infected_last) & (Infected_last != Symptom)) %>% mutate(band = Infected_last - Infected_first)
```

|  type |count   |
|---|---|
| upper bound only  |  29 |  
| lower bound only  | 133  | 
| lower and upper bound | 35 |
| knows the date of infection  |46   |
| total  |  243 |
For these patients we only have an lower bound for $T$.


```{r, echo=F}
par(mfrow = c(2,2))
hist(upper$upper, main = 'Upper bound only', xlab = 'max.incub')
hist(lower$lower, main = "Lower bound only", xlab = 'min.incub')
hist(exact$incub, xlab = 'exact incubation period', main = 'Exact infection date')
hist(band$band, xlab = 'max.incub - min.incub', main = 'Both upper and lower bound')
```


## Methods
We start with simple parametric models that does not incorporate covariates such as age and gender.

```{r}
symptom <- data$Symptom
infected_first <- data$Infected_first
infected_last <- data$Infected_last

loglikelihood <- function(par, family = 'gamma') {
    GT <- R0::generation.time(type = family, par, truncate = 100);
    loglike <- 0
    for (i in 1:nrow(data)) {
        min.incub <- symptom[i] - infected_last[i]
        max.incub <- symptom[i] - infected_first[i]
        if (max.incub == 0){
          loglike <- loglike + log(GT$GT[2])
        }else{
          loglike <- loglike + log(sum(GT$GT[1 + (min.incub):(max.incub)])) ## +1 for starting at zero
        }
        if (is.infinite(loglike)){
          print(i)
          print(data[i,])
          break
        }
    } 
    loglike 
}
```


```{r mle}
calculate_mle <- function(init = c(7.5, 3.5), family = 'gamma'){
  fit <- optim(par = init, fn = function(x) loglikelihood(x, family), control = list(fnscale = -1))
  fit
}
```


```{r}
parametric_families <- c('gamma','lognormal','weibull')
mle_table <- data.frame(distribution = parametric_families, mean = NA, sd = NA, loglikelihood = NA)
for (i in 1:length(parametric_families)){
  mle <- calculate_mle(family = parametric_families[i])
  mle_table[i, c(2,3)] <- as.numeric(mle$par)
  mle_table[i,4] <- as.numeric(mle$value)
}
print(mle_table)
```

The three parametric models have similar loglikelihood. We can look at the 95\% quantile for $T$ based on the gamma model. 
```{r quantile}
b <- mle_table[1,2] / mle_table[1,3] **2 
a <- mle_table[1,2] * b
## 95% quantile for T
c(qgamma(p = 0.025, shape = a, rate = b),qgamma(p = 1 - 0.025, shape = a, rate = b))
## IQR
c(qgamma(p = 0.25, shape = a, rate = b),qgamma(p = 1 - 0.25, shape = a, rate = b))
```

We can also look at the survival curve $P(T \geq t)$.
```{r survival}
t_v <- seq(0,30,1)
cmf <- pgamma(q = t_v, shape = a, rate = b)
plot(t_v, 1 - cmf, 
     xlab = 'time', ylab = 'survival probability', main = 'survival curve',
     type = 'l')
```


I choose to display the credible region based on the gamma model here.
```{r credible region}
pars <- expand.grid(mean = seq(5, 15, length.out = 100), sd = seq(3, 15, length.out = 100))
pars$loglike <- apply(pars, 1, FUN = function(par) loglikelihood(par))
# print(pars[which.max(pars$loglike),]) ## print the grid-search MLE to terminal
pars$in.CR <- (pars$loglike > mle_table[1,4] - qchisq(0.95, 1) / 2)
p2 <- ggplot(pars) + aes(x = mean, y = sd, fill = in.CR) +geom_tile()
p2
```

The incubation period estimate reported here is much higher than previous studies. This can partially be explained by the more up-to-date data that we collected. Earlier reports collected cases that are biased towards patients that display shorter incubation periods. We need to think through our subset selection criterion and examine if there is any bias. I also find some outliers in the data.

## Outliers 

Some cases have very large lower bound for $T_i$.
```{r outliers}
subset(data, Symptom - Infected_last > 20)
```

